---
title: "Teams"
date: 'Last updated: `r format(Sys.Date())`'
output:
  html_document:
    dpi: 500
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

Below you'll find figures showing the latest traditional and advanced metrics I'm tracking for all 30 MLB teams.

```{r, echo = FALSE}

library(baseballr)
library(tidyverse)
library(teamcolors) # team colors and logos
library(extrafont) # for extra fonts
library(ggrepel) # better labels
library(ggimage)
library(glue)
library(ggtext)
library(reactable)
library(reactablefmtr)
library(gt)
library(gtExtras)
library(webshot)
library(ggforce)
library(ggsci)
library(paletteer)
library(prismatic)

# Optional but makes R prefer not to display numbers in scientific notation
options(scipen = 9999)

# Custom ggplot theme (inspired by Owen Phillips at the F5 substack blog)
theme_custom <- function () { 
  theme_minimal(base_size=11, base_family="Chivo") %+replace% 
    theme(
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = 'floralwhite', color = "floralwhite")
    )
}

# Define an aspect ratio to use throughout. This value is the golden ratio which provides a wider than tall rectangle
asp_ratio <- 1.618 

# load teamcolors and logos
mlb_teamcolors <- read_csv('mlb_teamcolors.csv')


## 1. ---------- Pythagorean Win % Plot ----------------------

today <- Sys.Date()
al_standings <- bref_standings_on_date(today, "AL Overall", from = FALSE)
nl_standings <- bref_standings_on_date(today, "NL Overall", from = FALSE)

mlb_standings <- rbind(al_standings, nl_standings)

mlb_standings <- mlb_standings %>%
    arrange(desc(`W-L%`))

# join in team logos and colors
mlb_standings <- mlb_standings %>%
    left_join(mlb_teamcolors %>% 
                  select(name, team_abbr, division, primary, team_logo_espn), by = c("Tm" = "team_abbr"))

# clean up df  
mlb_standings <- mlb_standings %>%
    mutate(Rank = row_number()) %>%
    select(Rank, team_logo_espn, name, division, W:`pythW-L%`, primary)


# make visual
mlb_standings %>% 
    ggplot(aes(x = `W-L%`, y = `pythW-L%`)) + 
    geom_image(aes(image = team_logo_espn), size = 0.045, by = "width", asp = asp_ratio) +
    annotate("text", x=0.35, y=0.55, label = "Better than Actual Record", color = "#00BA38", family = 'Outfit') +
    annotate("text", x=0.55, y=0.35, label = "Worse than Actual Record", color = "#F8766D", family = 'Outfit') +
    theme_custom() +
    #geom_abline(slope = 1, alpha = .2) +
    geom_smooth(method = "lm", se = FALSE, color="gray", alpha = 0.2) +
    scale_y_continuous(labels = scales::percent_format()) +
    scale_x_continuous(labels = scales::percent_format()) +
    theme_custom() +
    labs(title = 'Pythagorean Wins',
         x = "Actual Win %",
         y = "Pythagorean Win %",
         subtitle = "Pythagorean Win % helps evaluate how teams perform vs expectations. Thru **{today}.**",
         caption = "Data: baseballR (Bill Petti) | Plot: @steodosescu") +
  theme(plot.title.position = "plot",
        plot.title = element_text(face = "bold", 
                                  size = 20, 
                                  hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 10,
      hjust = 0.5)
  ) +
  theme(plot.subtitle = element_markdown()) 
    

```

### Run Differential Evolution

```{r, echo = FALSE}

## 2. ---------- Run Differential Evolution Plot (WIP) ----------------------

# load MLB schedule
games <- baseballr::mlb_schedule(season = "2024") %>%
    filter(game_type == 'R')

# Process data frame to get one row per team-game (long format)
team_games <- games %>%
     select(teams_away_team_name, teams_away_score, teams_home_team_name, teams_home_score) %>%
    rename(away_team = teams_away_team_name,
         away_score = teams_away_score,
         home_team = teams_home_team_name,
         home_score = teams_home_score) %>%
    pivot_longer(
    cols = c(away_team, home_team),
    names_to = "home_away",
    values_to = "team"
  ) %>%
     mutate(score = ifelse(home_away == "home", home_score, away_score),
           opp_score = ifelse(home_away == "home", away_score, home_score
                              )
     )






```


### Performance vs Expectations

The below shows current win totals -- expressed in win totals -- compared to pre-season win total expectations, according to FanGraphs. Expected win total data was pulled during Spring Training in early March 2025.

```{r, echo = FALSE}

# load preseason win totals from Fangraphs
fangraphs <- read_csv("/Users/Stephan/Desktop/R Projects/MLB/fangraphs_projections_2025.csv")

fangraphs <- fangraphs %>% 
    left_join(mlb_teamcolors %>% 
                  select(name, team_abbr, mascot), by = c("team" = "mascot")) 
    
# join current standings data
expectations <- mlb_standings %>%
     left_join(fangraphs, by = c("name" = "name")) %>%
    mutate(win_total_diff = W - w)


# create barplot
expectations %>% 
  ggplot(aes(x = fct_reorder(team, -win_total_diff), y = win_total_diff)) +
  geom_col(aes(fill = primary,
               color = after_scale(clr_darken(fill, 0.3))
  ),
  width = 0.4, 
  alpha = .75,
  ) + 
  scale_color_identity(aesthetics =  c("fill"))  +
  geom_image(
    aes(
      image = team_logo_espn                                  
    ), 
    size = 0.025, 
    by = "width", 
    asp = asp_ratio
  ) +
  geom_hline(yintercept = 0, color = "black", size = 1) +
  scale_y_continuous(
      #labels = scales::percent_format(),
      breaks = scales::pretty_breaks(n = 10)) +
  theme_custom() + 
  theme(axis.text.x = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank()) + 
  labs(x = "", 
       y = "Wins over Expected", 
       caption = "Data: baseballR (Bill Petti/FanGraphs | Plot: @steodosescu",
       title = glue("Performance Relative to Expecations"),
       subtitle = glue("Actual win totals vs. Predicted. Estimated totals pulled from FanGraphs as of March 9, 2025.")) +
  theme(plot.title.position = "plot",
        plot.title = element_text(face = "bold", 
                                  size = 20, 
                                  hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5)
  )
  


```

