---
title: "Teams"
date: 'Last updated: `r format(Sys.Date())`'
output:
  html_document:
    dpi: 500
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

Below you'll find figures showing the latest traditional and advanced metrics I'm tracking for all 30 MLB teams.

```{r, echo = FALSE}

library(baseballr)
library(tidyverse)
library(teamcolors) # team colors and logos
library(extrafont) # for extra fonts
library(ggrepel) # better labels
library(ggimage)
library(glue)
library(ggtext)
library(ggforce)
library(ggsci)
library(paletteer)
library(prismatic)
library(gt)
library(gtExtras)

# Optional but makes R prefer not to display numbers in scientific notation
options(scipen = 9999)

# Custom ggplot theme (inspired by Owen Phillips at the F5 substack blog)
theme_custom <- function () { 
  theme_minimal(base_size=11, base_family="Chivo") %+replace% 
    theme(
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = 'floralwhite', color = "floralwhite"),
      plot.subtitle = element_text(color = "grey60")
    )
}

# Define an aspect ratio to use throughout. This value is the golden ratio which provides a wider than tall rectangle
asp_ratio <- 1.618 

# load teamcolors and logos
mlb_teamcolors <- read_csv('mlb_teamcolors.csv') %>% 
    mutate(team_abbr = case_when(team_abbr == "OAK" ~ "ATH", # A's are no longer called OAK
                     TRUE ~ team_abbr,)
           )
    


```


### Pythagorean Wins

The below shows each team's actual win percentage vs their pythagoroean win percentage. Pythagorean winning percentage is a formula developed by renowned statistician Bill James. The concept strives to determine the number of games that a team *should* have won -- based on its total number of runs scored versus its number of runs allowed. It can help identify teams that have either overachieved or underachieved, and better predict a team's future performance.

```{r Pythagorean Wins, echo = FALSE}

## 1. ---------- Pythagorean Win % Plot ----------------------

today <- Sys.Date()-1
al_standings <- bref_standings_on_date(today, "AL Overall", from = FALSE)
nl_standings <- bref_standings_on_date(today, "NL Overall", from = FALSE)

mlb_standings <- rbind(al_standings, nl_standings)

mlb_standings <- mlb_standings %>%
    arrange(desc(`W-L%`))

# join in team logos and colors
mlb_standings <- mlb_standings %>%
    left_join(mlb_teamcolors %>% 
                  select(name, team_abbr, division, primary, team_logo_espn), by = c("Tm" = "team_abbr"))

# clean up df  
mlb_standings <- mlb_standings %>%
    mutate(Rank = row_number()) %>%
    select(Rank, team_logo_espn, name, division, W:`pythW-L%`, primary)


# make visual
mlb_standings %>% 
    ggplot(aes(x = `W-L%`, y = `pythW-L%`)) + 
    geom_image(aes(image = team_logo_espn), size = 0.045, by = "width", asp = asp_ratio) +
    annotate("text", x=0.35, y=0.55, label = "Better than Actual Record", color = "#00BA38", family = 'Outfit') +
    annotate("text", x=0.55, y=0.35, label = "Worse than Actual Record", color = "#F8766D", family = 'Outfit') +
    theme_custom() +
    #geom_abline(slope = 1, alpha = .2) +
    geom_smooth(method = "lm", se = FALSE, color="gray", alpha = 0.2) +
    scale_y_continuous(labels = scales::percent_format()) +
    scale_x_continuous(labels = scales::percent_format()) +
    theme_custom() +
    labs(title = 'Pythagorean Wins',
         x = "Actual Win %",
         y = "Pythagorean Win %",
         subtitle = glue("Pythagorean Win % helps evaluate how teams perform vs expectations. Thru **{today}.**"),
         caption = "Data: baseballR (Bill Petti) | Plot: @steodosescu") +
  theme(plot.title.position = "plot",
        plot.title = element_text(face = "bold", 
                                  size = 20, 
                                  hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 10,
      hjust = 0.5)
  ) +
  theme(plot.subtitle = element_markdown()) 

```

***
### MLB Team Performance 

```{r RS vs RA, echo = FALSE}

## 2. ---------- Runs Scored vs Runs Allowed ----------------------

mlb_standings2 <- mlb_standings %>%
      mutate(gp = W+L,
           rs_g = RS/gp,
           ra_g = RA/gp)
    
mlb_standings2 %>% 
    mutate(gp = W+L,
           rs_g = RS/gp,
           ra_g = RA/gp) %>% 
    ggplot(aes(x = rs_g, y = ra_g)) + 
    geom_image(aes(image = team_logo_espn), size = 0.055, by = "width", asp = asp_ratio) +
    geom_hline(yintercept = mean(mlb_standings2$rs_g), color = "red", linetype = "dashed") +
    geom_vline(xintercept =  mean(mlb_standings2$ra_g), color = "red", linetype = "dashed") +
    theme_custom() +
  #geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3, -.4), alpha = .2) +
    labs(x = "Runs Scored/Game",
         y = "Runs Allowed/Game",
         caption = "Data: baseballR (Bill Petti) | Plot: @steodosescu",
         title = glue("MLB Scoring Performance"),
        subtitle = glue("2026 season, data thru {today}.")) +
  theme(plot.title.position = "plot",
        plot.title = element_text(face = "bold", 
                                  size = 20, 
                                  hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 10,
      hjust = 0.5)
  ) +
  theme(plot.subtitle = element_markdown()) +
  scale_y_reverse()

```


### Run Differential Evolution

Evolution of each team's run differential this season. 

```{r, Run Differential, echo = FALSE}

## 2. ---------- Run Differential Evolution Plot ----------------------

# load MLB schedule
games <- baseballr::mlb_schedule(season = "2026") %>%
    filter(game_type == 'R')

# Process data frame to get one row per team-game (long format)
team_games <- games %>%
     select(teams_away_team_name, teams_away_score, teams_home_team_name, teams_home_score) %>%
    rename(away_team = teams_away_team_name,
         away_score = teams_away_score,
         home_team = teams_home_team_name,
         home_score = teams_home_score) %>%
    pivot_longer(
    cols = c(away_team, home_team),
    names_to = "home_away",
    values_to = "team"
  ) %>%
     mutate(score = ifelse(home_away == "home_team", home_score, away_score),
           opp_score = ifelse(home_away == "home_team", away_score, home_score
                              )
     ) %>% 
    drop_na()


team_games2 <- team_games %>% 
    select(-c(home_score, away_score, home_away)) %>%
    group_by(team) %>%
    mutate(run_diff = score - opp_score) %>% 
    # calculate the running counts for run differential and games
    mutate(
           rd_running = cumsum(run_diff),
           game_count = row_number())

# join in teamcolors
team_games2 <- team_games2 %>% 
    left_join(mlb_teamcolors %>% 
                  select(name, team_abbr, primary, team_logo_espn), by = c("team" = "name")
              )

# determine endpoints for logos
endpoints <- team_games2 %>%
  group_by(team) %>%
  filter(!is.na(rd_running)) %>%
  slice_max(order_by = game_count, n = 1) %>%
  ungroup()


# make plot
team_games2 %>% 
    ggplot(aes(x=game_count, y = rd_running, group = team)) +
    geom_line(size = 1, color = "grey") +
    geom_image(data = endpoints, aes(x = game_count, y = rd_running, image = team_logo_espn), 
             size = 0.05, inherit.aes = FALSE) +
    labs(x = "", y = "Run Differential",
         title = "2026 MLB Run Differential",
         subtitle = glue("Daily cumulative run differential totals. Data thru {today} games."),
         caption = "Data: baseballR (Bill Petti) | Plot: @steodosescu",
         color = "") +
    scale_x_continuous(limits = c(1, 162)) +
    theme_custom() +
    #scale_color_manual(values = team_games2$primary) +
    theme(legend.position="none") +
    theme(plot.title.position = "plot",
        plot.title = element_text(face = "bold", 
                                  size = 20, 
                                  hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 10,
      hjust = 0.5)
  ) +
    theme(plot.title = element_text(face="bold")) +
    theme(plot.title = element_markdown()) +
    theme(plot.subtitle = element_markdown())




```


***
### Team ERA vs Team OPS
Comparison of team-level pitching (ERA) and hitting (OPS)

```{r ERA vs OPS, echo = FALSE}

# load Fangraphs stat categories
fg_team_batter <- fg_team_batter(startseason = 2025, endseason = 2025, qual = 400)
fg_team_pitcher <- fg_team_pitcher(startseason = 2025, endseason = 2025)
fg_team_fielder <- fg_team_fielder(startseason = 2025, endseason = 2025, qual = 400)

fg_team_batter <- fg_team_batter %>% 
    select(team_name, OBP, OPS, WAR, wOBA, wRC, XBR) %>% 
    rename(b_WAR = WAR)

fg_team_pitcher <- fg_team_pitcher %>% 
    select(team_name, WAR, ERA, AVG, WHIP) %>% 
    rename(p_WAR = WAR)

fg_team_fielder <- fg_team_fielder %>% 
    select(team_name, team_name_abb, FP, DRS)

# join with team colors
fg_totals <- fg_team_batter %>% 
    left_join(fg_team_pitcher) %>% 
    left_join(fg_team_fielder, by = c("team_name" = "team_name_abb")
              ) %>% 
    select(-team_name.y) %>% 
    left_join(mlb_teamcolors %>% 
                  select(name, team_abbr, division, primary, team_logo_espn), by = c("team_name" = "team_abbr")) %>% 
    mutate(WAR = b_WAR + p_WAR) %>% 
    select(-b_WAR, -p_WAR) %>% 
    arrange(desc(WAR))

# create OPS vs ERA chart
fg_totals %>% 
    ggplot(aes(x = ERA, y = OPS)) + 
    geom_image(aes(image = team_logo_espn), size = 0.055, by = "width", asp = asp_ratio) +
    geom_hline(yintercept = mean(fg_totals$OPS), color = "red", linetype = "dashed") +
    geom_vline(xintercept =  mean(fg_totals$ERA), color = "red", linetype = "dashed") +
    theme_custom() +
  #geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3, -.4), alpha = .2) +
    labs(x = "Team ERA",
         y = "Team OPS",
         caption = "Data: Fangraphs/baseballR (Bill Petti) | Plot: @steodosescu",
         title = glue("MLB Hitting vs Pitching Performance"),
        subtitle = glue("2025 season, data thru {today}.")) +
  theme(plot.title.position = "plot",
        plot.title = element_text(face = "bold", 
                                  size = 20, 
                                  hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 10,
      hjust = 0.5)
  ) +
  theme(plot.subtitle = element_markdown()) +
    scale_x_reverse(labels = scales::label_number(accuracy = 0.01))



```



***
### Performance vs Expectations

The below shows current win totals -- expressed in win totals -- compared to pre-season win total expectations, according to FanGraphs. Expected win total data was pulled during Spring Training in early March 2026.

```{r Expectations, echo = FALSE}

# load preseason win totals from Fangraphs
projections <- read_csv("fg_projections_2026.csv")

projections <- projections %>% 
    left_join(mlb_teamcolors %>% 
                  select(name, team_abbr, mascot), by = c("team" = "mascot")) 
    
# join current standings data
expectations <- mlb_standings %>%
     left_join(projections, by = c("name" = "name")) %>%
    mutate(gp = W+L,
        win_perc_actual= W/gp,
        win_total_diff = win_perc_actual - win_perc_avg)


# create barplot
expectations %>% 
  ggplot(aes(x = fct_reorder(team, -win_total_diff), y = win_total_diff)) +
  geom_col(aes(fill = primary,
               color = after_scale(clr_darken(fill, 0.3))
  ),
  width = 0.4, 
  alpha = .75,
  ) + 
  scale_color_identity(aesthetics =  c("fill"))  +
  geom_image(
    aes(
      image = team_logo_espn                                  
    ), 
    size = 0.025, 
    by = "width", 
    asp = asp_ratio
  ) +
  geom_hline(yintercept = 0, color = "black", size = 1) +
  scale_y_continuous(
      labels = scales::percent_format(),
      #breaks = scales::pretty_breaks(n = 10)
      ) +
  theme_custom() + 
  theme(axis.text.x = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank()) + 
  labs(x = "", 
       y = "Win % over Expected", 
       caption = "Data: baseballR (Bill Petti/FanGraphs/Odds API/Neil Paine/Clay Davenport | Plot: @steodosescu",
       title = glue("Performance Relative to Expecations"),
       subtitle = glue("Actual win % vs. preseason model and odds predictions. Estimated totals pulled from various source systems March 26, 2026.")) +
  theme(plot.title.position = "plot",
        plot.title = element_text(face = "bold", 
                                  size = 20, 
                                  hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5)
  )
  


```


***
### Attendance Profiles


```{r Attendance Leader, echo = FALSE}

# load attendance data
attendance <- baseballr::mlb_attendance(league_list_id = 'mlb', season = 2026)

# Get the team with the highest average attendance
attendance_leader <- attendance %>%
  filter(!is.na(attendance_average_home)) %>%
  slice_max(order_by = attendance_average_home, n = 1) %>%
  mutate(
    leader_line = glue("The **{team_name}** lead the league in attendance, averaging **{scales::comma(attendance_average_home)}** fans per game.")
  )

```

2026 Attendance totals. `r attendance_leader$leader_line`

```{r Attendance, echo = FALSE}

attendance <- attendance %>% 
    left_join(mlb_teamcolors %>% 
                  select(team_id, name, team_abbr, primary, team_logo_espn), by = "team_id")

# create attendance barplot
attendance %>% 
    drop_na() %>%
  ggplot(aes(x = fct_reorder(name, -attendance_average_home), y = attendance_average_home)) +
  geom_col(aes(fill = primary,
               color = after_scale(clr_darken(fill, 0.3))
  ),
  width = 0.4, 
  alpha = .75,
  ) + 
  scale_color_identity(aesthetics =  c("fill"))  +
  geom_image(
    aes(
      image = team_logo_espn                                  
    ), 
    size = 0.025, 
    by = "width", 
    asp = asp_ratio
  ) +
  geom_hline(yintercept = 0, color = "black", size = 1) +
    scale_y_reverse() +
  scale_y_continuous(
      labels = scales::label_number(scale_cut = scales::cut_short_scale())
      ) +
  theme_custom() + 
  theme(axis.text.x = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank()) + 
  labs(x = "", 
       y = "Average Home Attendance", 
       caption = "Data: baseballR (Bill Petti) | Plot: @steodosescu",
       title = glue("Take me out to the Ballgame"),
       subtitle = glue("Average home attendance totals in 2026. Thru **{today}.**")) +
  theme(plot.title.position = "plot",
        plot.title = element_text(face = "bold", 
                                  size = 20, 
                                  hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5)
  ) +
  theme(plot.subtitle = element_markdown()) 
  


```


***
### Team Ranks

League rankings among several popular advanced stats metrics. 

```{r Team Ranks, echo = FALSE}

# create ranking columns for each stat
fg_ranks <- fg_totals %>%
  mutate(across(c(OBP, OPS, WAR, wOBA, wRC, XBR, FP, DRS), ~ min_rank(desc(.x)), .names = "{.col}_rank"),
         across(c(ERA, WHIP, AVG), ~ min_rank(.x), .names = "{.col}_rank")
  )

# create GT table
fg_ranks %>% 
    mutate(rank = row_number()) %>% 
    select(rank, team_logo_espn, team_name, OBP_rank:AVG_rank, -WAR_rank) %>% 
    gt() %>% 
    cols_label(rank = "",
               team_logo_espn = "",
               team_name = "Team",
               OBP_rank = "OBP",
               OPS_rank = "OPS",
              # WAR_rank = "WAR",
               wOBA_rank = "wOBA",
               wRC_rank = "wRC",
               XBR_rank = "XBR",
               FP_rank = "FP",
               DRS_rank = "DRS",
               ERA_rank = "ERA",
              WHIP_rank = "WHIP",
               AVG_rank = "AVG"
               ) %>% 
  #    cols_width(
  #        everything() ~ px(60)
  # ) %>% 
     tab_spanner(
    label = "Batting/Baserunning",
    columns = c(4:8)
  ) %>%
     tab_spanner(
    label = "Fielding",
    columns = c(9:10)
  ) %>%
     tab_spanner(
    label = "Pitching",
    columns = c(11:13)
  ) %>%
     gt_img_rows(columns = team_logo_espn, img_source = "web", height = 30) %>%
     data_color(
    columns = 4:9, 
    colors = scales::col_numeric(
      palette = paletteer::paletteer_d(
        palette = "ggsci::amber_material",
        direction = -1 # Offensive stats
      ) %>% as.character(),
      domain = NULL, 
      na.color = "#005C55FF"
    )) %>%
     data_color(
    columns = 10:13, 
    colors = scales::col_numeric(
      palette = paletteer::paletteer_d(
        palette = "ggsci::amber_material",
        direction = 1 #Pitching stats
      ) %>% as.character(),
      domain = NULL, 
      na.color = "#005C55FF"
    )) %>%
  cols_align(align = "left",
             columns = 1) %>%
  tab_header(title = md("**MLB Team Ranks**"),
             subtitle = glue("League rankings (1-30) across several advanced statistical categories sourced from MLB, Fangraphs, and Statcast data. Sorted by overall team WAR (Fangraphs). Data thru games as of {today}.")) %>%
  tab_source_note(
    source_note = md("DATA: baseballR (Bill Petti)<br>TABLE: @steodosescu")) %>% 
    tab_options(
        column_labels.background.color = "white",
        table.border.top.width = px(3),
        table.border.top.color = "white",
        table.border.bottom.color = "white",
        table.border.bottom.width = px(3),
        column_labels.border.top.width = px(3),
        column_labels.border.top.color = "white",
        column_labels.border.bottom.width = px(3),
        column_labels.border.bottom.color = "black",
        data_row.padding = px(3),
        source_notes.font.size = 12,
        table.font.size = 16,
        heading.align = "center"
    ) %>% 
    tab_style(
        style = cell_text(weight = "bold"),
        locations = cells_column_labels(everything())
        ) 
# %>%
#     gtsave('MLB Stats Table.png')


```

